\begin{abstract}
Sign language is the primary mode of communication for millions of deaf and
hard-of-hearing individuals worldwide. While American Sign Language (ASL)
recognition has been studied extensively, Arabic Sign Language (ArSL) remains
underexplored despite serving an estimated seven million Arabic-speaking deaf
individuals. In this work, we present ArSignTutor, an interactive real-time
learning system for the Arabic Sign Language alphabet. Our approach combines a
fine-tuned MobileNetV2 convolutional neural network with MediaPipe-based hand
detection for robust, real-time gesture classification, and pairs it with a
spelling-based tutoring interface that guides users through word formation
letter-by-letter with immediate visual feedback. Training MobileNetV2 on the
ArASL dataset (over 54{,}000 images) under a random image-level split yields
99.26\% test accuracy; we show, however, that this figure is inflated by
\emph{data leakage}, since the dataset contains many near-duplicate frames of the
same signer that a random split distributes across both training and test sets. To
obtain an honest estimate of generalization, we evaluate the model across datasets
on an independent RGB Arabic alphabet dataset (31 shared classes), where accuracy
falls to 48.78\%, and then retrain under a leakage-aware (cluster-based) split with
hand-region preprocessing, reaching a defensible 94.33\% test accuracy. We further
introduce a selective hand-aware background-blurring technique to improve the
clarity of the real-time interface. Our main contribution is methodological: a
reproducible protocol that separates genuine generalization from leakage-driven
optimism, alongside a practical interactive tool for the Arabic-speaking deaf
community.
\end{abstract}